Orienting the causal relationship between imprecisely measured traits using GWAS summary data

### 2017-10-05 docs/manuscript.tex
split revision into supplement and main texts

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{27 July 2017}

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population (41). It is calculated as

@@

As we show in Appendix 2 it is possible to infer the bounds of
measurement error on \(x_o\) or \(y_o\) given known genetic
associations. The maximum measurement imprecision of \(x_o\) is
\(\rho_{g,x_o}\), because it is known that at least that much of the
variance has been explained in \(x_o\) by \(g\). The minimum is 0,
denoting perfectly measured trait values (the same logic applies to
\(y_o\)). It is possible to simulate what the inferred causal direction
would be for all values within these bounds.

To evaluate how reliable, \(R\), the inference of the causal direction
is to potential measurement error in \(x\) and \(y\) we need to predict
the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
error. We offer two tools in which to do this. First, the user can
provide values of measurement error for \(x\) and \(y\) and obtain a
revised inference of the causal direction. Second, we integrate over the
entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
measurement error values, assuming that any measurement error value is
equally likely. Across all possible values of measurement error in \(x\)
and \(y\) we find the volume that agrees with the inferred direction of
causality and the volume that disagrees with the inferred direction of
causality, and take the ratio of these two values. A ratio \(R=1\)
indicates that the inferred causal direction is highly sensitive to
measurement error, because equal weight of the measurement error
parameter space supporting both directions of causality. In general, the
\(R\) value denotes that the inferred direction of causality is \(R\)
times more likely to be the empirical result than the opposite
direction. Full details are provided in Appendix 2.

@@

In the causal inference test (CIT), the 4th condition (see Methods)
employs mediation for causal inference, and can be expressed as
\(cov(g, y - \hat{y}) = 0\), where
\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement error
in scale and imprecision is introduced, such that \(y_o\) is the
measured value of \(y\), it can be shown using basic covariance
properties (Appendix 1) that

@@

For a particular association, it is of interest to identify the range of
possible measurement error values for which the method will give results
that agree or disagree with the empirically inferred causal direction
(Figure 4a, Appendix 2). This metric can be used to evaluate the
reliability of MR Steiger test.

We show that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2) determines
the range of parameters around which the MR Steiger test is liable to
provide the wrong direction of causality (\emph{i.e.} if \(d>0\) then
the MR Steiger test is likely to be correct about the causal direction).
Figure 4b shows that when there is no measurement error in \(x\), the MR
Steiger test is unlikely to infer the wrong direction of causality even
if there is measurement error in \(y\). It also shows that in most cases
where \(x\) is measured with error, especially when the causal effect
between \(x\) and \(y\) is not very large, the sensitivity of the MR
Steiger test to measurement error is relatively low.

Unmeasured confounding between the exposure and outcome can also give
rise to problems with the MR Steiger approach (Appendix 3). The
relationship between unmeasured confounding and causal orientation is
complex across the parameter space of possible confounding values
(Supplementary Figure 1). Based on the range of parameter values that we
explored, when the magnitude of the observational variance explained
between the exposure and the outcome is below 0.2 the MR Steiger method
is unlikely to return the incorrect causal direction due to unmeasured
confounding.

@@

\subsection{Acknowledgements}\label{acknowledgements}

This work was funded by the UK Medical Research Council (MC\_UU\_12013/1
and MC\_UU\_12013/9).

@@

\subsection{Appendix 1}\label{appendix-1}

We assume the following model

\[
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
\end{aligned}
\]

where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
that has a direct effect on \(x\), \(x_o\) is the measured quantity of
\(x\), where measurement error is incurred from linear transformation in
\(\alpha_{mx}\) and \(\beta_{mx}\) and imprecision from
\(\epsilon_{mx}\), and \(y_o\) is the measured quantity of \(y\), where
measurement error is incurred from linear transformation in
\(\alpha_{my}\) and \(\beta_{my}\) and imprecision from
\(\epsilon_{my}\). Our objective is to estimate the expected magnitude
of association between \(g\) and \(y\) after conditioning on \(x\).
Under the CIT, this is expected to be \(cov(g, y_o - \hat{y}_o) = 0\)
when \(x\) causes \(y\), where
\(\hat{y}_o = \hat{a}_{x_o} + \hat{\beta}_{x_o} x_o\) is the predicted
value of \(y_o\) using the measured value of \(x_o\).

We can split \(cov(g, y_o - \hat{y}_o)\) into two parts, \(cov(g, y_o)\)
and \(cov(g, \hat{y}_o)\).

\textbf{Part 1}

\[
\begin{aligned}
cov(g, y_o) & = cov(g, \beta_{my} y) \\
            & = cov(g, \beta_{my} \beta_x x) \\
            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
            & = \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

\textbf{Part 2}

\[
\begin{aligned}
cov(g, \hat{y}_o) & = cov(g, \hat{\beta}_{x_o} x_o) \\
                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} x) \\
                  & = cov(g, \hat{\beta}_{x_o} \beta_{mx} \beta_g g) \\
                  & = \hat{\beta}_{x_o} \beta_{mx} \beta_g var(g)
\end{aligned}
\]

Simpifying further

\[
\begin{aligned}
\hat{\beta}_{x_o} & = \frac{cov(y_o, x_o)} {var(x_o)} \\
                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
\end{aligned}
\]

which can be substituted back to give

\[
\begin{aligned}
cov(g, \hat{y}_o) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

Finally

\[
\begin{aligned}
cov(g, y_o - \hat{y}_o) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
\]

thus \(cov(g, y_o - \hat{y}_o) = 0\) if the measurement imprecision in
\(x_o\) is \(var(\epsilon_m) = 0\). However if there is any imprecision
then the condition \(cov(g, y_o - \hat{y}_o) = 0\) will not hold.

\subsection{Appendix 2}\label{appendix-2}

Assuming that either \(x \rightarrow y\) or \(y \rightarrow x\), the
causal direction can be inferred by evaluating which of \(\rho_{g, x}\)
and \(\rho_{g, y}\) is larger in magnitude. The Steiger test is a
hypothesis test that provides a p-value for observing the difference in
these correlations under the null hypothesis that they are equal.

Assuming the causal direction is \(x \to y\), two stage MR is formulated
using the following regression models:

\[
x = \alpha_1 + \beta_1 g + e_1
\]

for the first stage and

\[
y = \alpha_2 + \beta_2 \hat{x} + e_2
\]

where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
the exposure variable \(x\), and it is expected that
\(\rho_{g, x} > \rho_{g, y}\) because
\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
causal association between \(x\) and \(y\) (which is likely to be less
than 1).

In the presence of measurement error in \(x\) and \(y\), however, the
empirical inference of the causal direction will instead be based on
evaluating \(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:

\[
\begin{aligned}
\rho_{g, x_O} & > \rho_{g, y_O} \\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
\end{aligned}
\]

In order to assess how reliable the inference of the causal direction is
in the presence of measurement imprecision, we can evaluate the range of
potential values of measurement error in \(x\) and \(y\) over which the
empirical difference in \(\rho_{g, x_o}\) and \(\rho_{g, y_o}\) would
return the wrong causal direction.

For different values of \(\rho_{x,x_o}\),
\(\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}\) and
\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\). For different values of
\(\rho_{y,y_o}\), \(\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}\)
and \(\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1\).

Call \(z = \rho_{g,y} - \rho_{g,x}\) the true difference in the variance
explained by the genetic variant in \(y\) and \(x\). If \(z < 0\) then
we infer that \(x \rightarrow y\). There will be some values of
\(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) that do not alter whether
\(z < 0\). To evaluate the reliability, \(R\), of the inference of the
causal direction with regards to measurement error, the objective is to
compare the proportion of the parameter space that agrees with the
inferred direction against the proportion which does not:

\[
R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
\]

If \(R=1\) then the direction of causality is equally probable across
the range of possible measurement error values. If \(R > 1\) then \(R\)
times as much of the parameter space favours the inferred direction of
causality. \(V_{z}\), the total volume of the function (Figure 4), can
be obtained analytically by solving:

\[
\begin{aligned}
V_z & = \int^1_{\rho_{g,x_o}} \int^1_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = \rho_{g,x_o}log(\rho_{g,x_o}) - \rho_{g,y_o}log(\rho_{g,y_o}) + \rho_{g,x_o}\rho_{g,y_o}(log(\rho_{g,y_o})-log(\rho_{g,x_o}))
\end{aligned}
\]

\(V_{z \ge 0}\), the proportion of the volume that lies above the
\(z=0\) plane, can also be obtained analytically. The region of this
volume is bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\)
where \(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,

\[
\begin{aligned}
V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
\end{aligned}
\]

Thus \(V_{z < 0} = V_{z} - V_{z \geq 0}\).

@@

\subsection{Appendix 3}\label{appendix-3}

We have assumed no unmeasured confounding in these simulations.
Unmeasured confounding will however have potentially large influences on
mediation-based methods for inferring causal directions, and can also
adversely influence the estimate of the causal direction for the Steiger
test.

\subsubsection{Unmeasured confounding in
mediation}\label{unmeasured-confounding-in-mediation}

Including an unmeasured confounder, \(u\), after ignoring intercept
terms the exposure \(x\) and outcome \(y\) variables can be modelled as

\[
\begin{aligned}
y & = \beta_x x + \beta_{uy} u + \epsilon_x \\
x & = \beta_g g + \beta_{ux} u + \epsilon_g
\end{aligned}
\]

The observational estimate of the causal effect of \(x\) on \(y\),
\(\hat{\beta}_x\) is obtained from

\[
\begin{aligned}
\hat{\beta}_x & = cov(x, y) / var(x) \\
& = \frac{\beta_g^2 \beta_x var(g) + \beta_{ux}^2 \beta_x var(u) + \beta_x var(\epsilon_g)} {\beta_g^2 var(g) + \beta_{ux}^2 var(u) + var(\epsilon_g)}
\end{aligned}
\]

From this it is clear that \(\beta_x\) and \(\hat{\beta}_x\) will differ
when both \(\beta_{uy}\) and \(\beta_{ux}\) are non-zero. Relating to
mediation, where we attempt to test if \(g\) associates with \(y\) after
adjusting \(y\) for \(x\), such that

\[
\hat{y}^* = \hat{\beta}_x x
\]

\[
\begin{aligned}
cov(g, y - \hat{y}^*) & = cov(g, \beta_x x + \beta_{uy} u + \epsilon_x - \hat{\beta}_x x) \\
& = cov(g, (\beta_x - \hat{\beta}_x)(\beta_g g + \beta_ux u + \epsilon_x)) \\
& = (\beta_x - \hat{\beta}_x) var(g)
\end{aligned}
\]

should any amount of unmeasured confounding exist, therefore, there will
remain an association between \(g\) and \(y|x\), which will introduce
errors in inferring causal directions.

\subsubsection{Unmeasured confounding in the MR Steiger
test}\label{unmeasured-confounding-in-the-mr-steiger-test}

Similarly, we can investigate the extent to which unmeasured confounding
will lead to the wrong causal direction between \(x\) and \(y\) using
the MR Steiger test, evaluating the liability for the inequality
\(cor(g,x)^2 > cor(g,y)^2\) being incorrect. After some algebra

\[
cor(g,x)^2 = \frac{\beta_g^2}{\beta_g^2var(g) + \beta_{ux}^2 var(u) + var(\epsilon_x)}
\]

\[
cor(g,y)^2 = \frac{\beta_x^2 \beta_g^2 var(g)^2} {\hat{\beta}_x^2 \beta_g^2 var(g) + \hat{\beta}_x^2 \beta_{ux}^2 var(u) + \beta_{uy}^2 var(u) + var(\epsilon_y)}
\]

Supplementary Figure 1 shows the relationship between the magnitude of
the correlations between \(x\), \(y\) and the confounder \(u\) for a
range of \(\beta_{xy} = (-2,2)\), \(\beta_{gx} = 0.1\) and a range of
confounder effects. The pattern of results were similar for different
values of \(\beta_{gx}\). We note that in most cases for the parameter
values explored, where the observational absolute \(\hat{R}^2_{xy}\) is
less than 0.2, unmeasured confounding will not incur the wrong causal
direction in the MR Steiger test.

\includegraphics{../images/unmeasured_conf_steiger.png}

Supplementary Figure 1: Graph representing the unmeasured confounding
parameters that will lead to the MR Steiger test returning the wrong
causal direction. Columns of boxes represent different signed values of
the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).

>>>>

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using GWAS summary data}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{28 July 2017}

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Supplementary
text 2), formalised by testing for a difference in the correlations
\(\rho_{gx}\) and \(\rho_{gy}\) using Steiger's Z-test for correlated
correlations within a population (41). It is calculated as

@@

As we show in Supplementary text 2 it is possible to infer the bounds of
measurement error on \(x_o\) or \(y_o\) given known genetic
associations. The maximum measurement imprecision of \(x_o\) is
\(\rho_{g,x_o}\), because it is known that at least that much of the
variance has been explained in \(x_o\) by \(g\). The minimum is 0,
denoting perfectly measured trait values (the same logic applies to
\(y_o\)). It is possible to simulate what the inferred causal direction
would be for all values within these bounds.

To evaluate how reliable, \(R\), the inference of the causal direction
is to potential measurement error in \(x\) and \(y\) we need to predict
the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
error. We offer two tools in which to do this. First, the user can
provide values of measurement error for \(x\) and \(y\) and obtain a
revised inference of the causal direction. Second, we integrate over the
entire range of \(\rho_{gy} - \rho_{gx}\) values for possible
measurement error values, assuming that any measurement error value is
equally likely. Across all possible values of measurement error in \(x\)
and \(y\) we find the volume that agrees with the inferred direction of
causality and the volume that disagrees with the inferred direction of
causality, and take the ratio of these two values. A ratio \(R=1\)
indicates that the inferred causal direction is highly sensitive to
measurement error, because equal weight of the measurement error
parameter space supporting both directions of causality. In general, the
\(R\) value denotes that the inferred direction of causality is \(R\)
times more likely to be the empirical result than the opposite
direction. Full details are provided in Supplementary text 2.

@@

In the causal inference test (CIT), the 4th condition (see Methods)
employs mediation for causal inference, and can be expressed as
\(cov(g, y - \hat{y}) = 0\), where
\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement error
in scale and imprecision is introduced, such that \(y_o\) is the
measured value of \(y\), it can be shown using basic covariance
properties (Supplementary text 1) that

@@

For a particular association, it is of interest to identify the range of
possible measurement error values for which the method will give results
that agree or disagree with the empirically inferred causal direction
(Figure 4a, Supplementary text 2). This metric can be used to evaluate
the reliability of MR Steiger test.

We show that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Supplementary text 2)
determines the range of parameters around which the MR Steiger test is
liable to provide the wrong direction of causality (\emph{i.e.} if
\(d>0\) then the MR Steiger test is likely to be correct about the
causal direction). Figure 4b shows that when there is no measurement
error in \(x\), the MR Steiger test is unlikely to infer the wrong
direction of causality even if there is measurement error in \(y\). It
also shows that in most cases where \(x\) is measured with error,
especially when the causal effect between \(x\) and \(y\) is not very
large, the sensitivity of the MR Steiger test to measurement error is
relatively low.

Unmeasured confounding between the exposure and outcome can also give
rise to problems with the MR Steiger approach (Supplementary text 3).
The relationship between unmeasured confounding and causal orientation
is complex across the parameter space of possible confounding values
(Supplementary figure 1). Based on the range of parameter values that we
explored, when the magnitude of the observational variance explained
between the exposure and the outcome is below 0.2 the MR Steiger method
is unlikely to return the incorrect causal direction due to unmeasured
confounding.

@@

@@

@@

\subsection{Supplementary information}\label{supplementary-information}

\textbf{Supplementary text 1} The influence of measurement error in the
exposure on mediation-based estimated

\textbf{Supplementary text 2} Sensitivity analysis for measurement error
on the MR Steiger test

\textbf{Supplementary text 3} The influence of unmeasured confounding on
the inference of causal directions

\textbf{Supplementary figure 1} Graph representing the unmeasured
confounding parameters that will lead to the MR Steiger test returning
the wrong causal direction. Columns of boxes represent different signed
values of the observational variance explained between \(x\) and \(y\)
(\(R^2_{xy}\)).
